Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-05-01 SupportingInformation.Rmd
edits, mostly for new holder materials

\noindent The matrices from which root character states and changes 
across phylogeny branches were randomly chosen were generated as follows.
For extreme right-tail dependent noise, we first produced
a $1000000$ by $2$ matrix, $m$, of 
independent standard-normal random draws. Then, each row of $m$ 
was identified for a first set of modifications randomly 
and independently with a $50\%$ chance; 
for those rows identified, every element of the row was replaced with the 
absolute value of the first element of the row (including the first element itself).
For other rows, every element in the row was replaced by the negative of its
own absolute value. This produces right-tail dependent noise, normally distributed
for each location. If left-tail dependent noise was desired, we took the negative of the
entire matrix. The moderately left-tail dependent matrix was based on a C copula 
with parameter chosen (using the `iRho` function in the `copula` package) so that
the Spearman correlation would be the same as for the extreme tail dependence 
datasets. The moderately right-tail dependent matrix and the symmetric tail dependence
matrix were constructed similarly, but using an SC copula and a N copula,
respectively.

Once character evolution was simulated 100 times for each dependence structure, 
we had 817 bivariate characters for each of 500 simulations. 
<!--***DAN: Shya to insert as the rest of this paragraph a description of what 
you did to each of these datasets to assess symmetry or asymmetry of tail dependence-->

<!--***DAN: Shya, please insert a new para here with a description of results, including a new figure with 5 panels showing the 5 histograms you showed me.-->

<!--***DAN: Dan does not understand the following two paragraphs yet, but is planning to
speak to Mark about them-->
Performing statistical tests for associations between continuous traits in different species
    was a primary motivating example of phylogenetic comparative methods.
For example, Felsenstein's method of phylogenetically independent contrasts 
    [@Felsenstein1985] is currently the second most-cited paper in the history
    of the journal the American Naturalist [@HueyGT2019].
Frustratingly, the field still lacks a dependable procedure for dealing with branch lengths, 
    which are a crucial input to the method.
Ideally, the branch lengths used to correct for phylogenetic effects would represent
    the expected amounts of change for the characters that are being analyzed.
Because researchers almost never have a reliable method for providing such branch lengths,
    most researchers rely on ultrametric trees -- those for which the branch
    length can be treated as a proxy for the duration of the branch in time.
Frequently these branch lengths are transformed to assess sensitivity to different 
    assumptions about the degree of phylogenetic inertia displayed by the traits
    under study [@Ives2018;@Harmon2018].
Even if one were able to simply use a time-based set of branch lengths, assigning
    dates to nodes in phylogenies is difficult.
DNA sequence data can provide estimates of branch lengths, but these estimates
    are dependent on the adequacy of models which correct sequences for multiple
    substitutions occurring at the same location.
Biases in estimating the evolutionary distance can affect down stream analyses [@Phillips2009].
Additionally, changes
    in the rate of molecular evolution make the estimation of dates difficult [@HeathM2014] even when branch lengths in time are accurately estimated.

Without reliable branch length estimates, it is difficult to interpret the significance
    of the magnitude of changes in traits across a tree.
Developing tests of association based on copula structure may possibly lead the way to more
    robust methods for studying associations when we lack defensible estimates of
    branch lengths. 
We note that our phylogenetic analyses here consisted merely of simulations to assess
    whether interesting copula structure in a simple evolutionary process could 
    leave a detectable signal on the data.
Substantial work remains to be done before we have a copula-based method for 
    analyzing data on a phylogenetic tree. 
But the nonparametric nature of approaches
    based solely on data ranks, as some of our copula methods are, may be a promising avenue for 
    avoiding inaccuracies that arise via the highly structured model assumptions implicit
    in the method of phylogenetically independent contrasts and related methods.

@@

>>>>

\noindent The matrices from which root character states and changes 
across phylogeny branches were randomly chosen were generated as follows.
For extreme right-tail dependent noise, we first produced
a $1000000$ by $2$ matrix, $m$, of 
independent standard-normal random draws. Then, each row of $m$ 
was identified for a first set of modifications randomly 
and independently with a $50\%$ chance; 
for those rows identified, every element of the row was replaced with the 
absolute value of the first element of the row (including the first element itself).
For other rows, every element in the row was replaced by the negative of its
own absolute value. This produces right-tail dependent noise, normally distributed
for each location. If left-tail dependent noise was desired, we took the negative of the
entire matrix. The moderately left-tail dependent matrix was based on a C copula 
with parameter chosen (using the `iRho` function in the `copula` package) so that
the Spearman correlation would be the same as for the extreme tail dependence 
datasets. The moderately right-tail dependent matrix and the symmetric tail dependence
matrix were constructed similarly, but using an SC copula and an N copula,
respectively.

Once character evolution was simulated 100 times for each dependence structure, 
we had 817 bivariate characters for each of 500 simulations. We thus analyzed 500 corresponding 
bivariate copulas (one for each simulation), each comprised of 817 points. 
To assess the symmetry or asymmetry of tail dependence for each copula, we computed 
the three non-parametric statistics $\cor_{0,0.2}-\cor{0.8,1}$, $\Ps_{0,0.2}-\Ps{0.8,1}$,
and $\Dsq_{0.8,1}-\Dsq_{0,0.2}$ for each copula. For each statistic and for each
dependence structure, we made a histogram (Fig. \ref{fig_cop_evo}). Values of these statistics less than 0
indicate lower-tail dependence in characters of extant species, whereas values greater
than 0 indicate upper-tail dependence. Asymmetry of tail dependence in evolutionary changes to characters
was associated with similar asymmetry of tail dependence in extant characters.

Performing statistical tests for associations between continuous traits in different species
    was a primary motivating example of phylogenetic comparative methods.
For example, Felsenstein's method of phylogenetically independent contrasts 
    [@Felsenstein1985] is currently the second most-cited paper in the history
    of the journal the American Naturalist [@HueyGT2019].
Frustratingly, the field still lacks a dependable procedure for dealing with branch lengths, 
    which are a crucial input to the method.
Ideally, the branch lengths used to correct for phylogenetic effects would represent
    the expected amounts of change for the characters that are being analyzed.
Because researchers almost never have a reliable method for providing such branch lengths,
    most researchers rely on ultrametric trees -- those for which the branch
    length can be treated as a proxy for the duration of the branch in time.
Frequently these branch lengths are transformed to assess sensitivity to different 
    assumptions about the degree of phylogenetic inertia displayed by the traits
    under study [@Ives2018;@Harmon2018].
Even if one were able to simply use a time-based set of branch lengths, assigning
    dates to nodes in phylogenies is difficult.
DNA sequence data can provide estimates of branch lengths, but these estimates
    are dependent on the adequacy of models which correct sequences for multiple
    substitutions occurring at the same location.
Biases in estimating the evolutionary distance can affect down stream analyses [@Phillips2009].
Additionally, changes
    in the rate of molecular evolution make the estimation of dates difficult [@HeathM2014] even when branch lengths in time are accurately estimated.

Without reliable branch length estimates, it is difficult to interpret the significance
    of the magnitude of changes in traits across a tree.
Developing tests of association based on copula structure may possibly lead the way to more
    robust methods for studying associations when we lack defensible estimates of
    branch lengths. 
We note that our phylogenetic analyses here consisted merely of simulations to assess
    whether interesting copula structure in a simple evolutionary process could 
    leave a detectable signal on the trait data for extant species.
Substantial work remains to be done before we have a copula-based method for 
    analyzing data on a phylogenetic tree. 
But the nonparametric nature of approaches
    based solely on data ranks, as some of our copula methods are, may be a promising avenue for 
    avoiding inaccuracies that arise via the highly structured model assumptions implicit
    in the method of phylogenetically independent contrasts and related methods.

@@

<!--Following chunk is to examine the hypothesis for third probable mechanism : 
asymmetric tail dependence in evolutionary changes in bivariate characters can give
rise to similarly asymmetric tail dependence in character distributions across phylogeny tips-->

```{r mechanism_evo_cop, echo=F, results="hide", cache=T, cache.extra=list(seed,mtime("explore_evo_cop.R"),mtime("CopulaFunctions_flexible.R"))}
set.seed(seed)

source("explore_evo_cop.R")
```

<!--adding plots showing results to confirm hypothesis behind third mechanism-->
\begin{figure}[!h]
\textbf{ \hspace{6.5 cm} (A) \hspace{2 cm} (B) \hspace{2 cm} (C)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/ExtremeL_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (D) \hspace{2 cm} (E) \hspace{2 cm} (F)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/SomewhatL_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (G) \hspace{2 cm} (H) \hspace{2 cm} (I)} \\
\vspace{-0.7 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/BivNormal_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (J) \hspace{2 cm} (K) \hspace{2 cm} (L)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/SomewhatR_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (M) \hspace{2 cm} (N) \hspace{2 cm} (O)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/ExtremeR_hist_taildep.pdf}\\
\caption[Non-parametric statistics on a mechanism of tail dependence in character distributions]{Three 
measures of asymmetry of tail dependence between two characters, across extant species ($\cor_{0,0.2}-\cor_{0.8,1}$ (A, D, G, J, M), 
$\Ps_{0,0.2}-\Ps_{0.8,1}$ (B, E, H, K, N), and $\Dsq_{0.8,1}-\Dsq_{0,0.2}$ (C, F, I, L, O)) were 
computed for each of 100 simulations for each of five types of tail dependence between 
evolutionary changes in the characters (extreme left-tail dependence (A-C), moderate left-tail dependence (D-F), 
symmetry of tail dependence (G-I), moderate right-tail dependence (J-L), and extreme right-tail dependence 
(M-O)). Asymmetry of tail dependence in evolutionary changes in characters tended to beget similar
asymmetry in the characters across extant species. For additional details see 
Appendix \ref{evol_mech}.\label{fig_cop_evo}}
\end{center}
\end{figure}
